---
title: "BFH_comment_charts"
author: Sophie Moullin
output:
  html_document: default
  word_document: default
---

# code for the graphics in Moullin and Harkness (2021, AJS)
# uses code examples from Healy's Data Vizualization (PUP), see https://socviz.co/

# uses the data tables produced from LISSY data. LIS data requires Stata code, and that it be run through those who house the data. This code was authored by Susan Harkness, and is available as an online appendix to the article. 

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(openxlsx)
library(here) 
```

```{r fig 1, echo=FALSE}
#Figure 1: Prevalances
LIS_prev <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Prevalence") 
colnames(LIS_prev) <- gsub("\\.", " ", colnames(LIS_prev)) # prevents conversion of space to dot

# group prevalences by country and order by prevalences
LIS_prev <- LIS_prev %>% 
          gather(household, prevalence, -(Country)) %>%  mutate(Country=fct_reorder(Country, desc(prevalence), na.rm=T))
View(LIS_prev) 

prev <- ggplot(data = LIS_prev,
               aes(x=Country, y=prevalence, color=household, na.rm=F))
prevalence <- prev + geom_point(na.rm = F) + labs(x=NULL, y="% of households with head < 65", title="Prevalence of Single Mother Households",
                                          caption="Source: Luxembourg Income Study.")+ theme(legend.position = "top", legend.title=element_blank(), axis.text.x = element_text(angle=45)) + scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(0,0.125)) 
prevalence
ggsave(here("Figures", "prevalence.pdf"))
```

# Figure 1: Prevalence of Single Mother Headed Households and Households Containing Single Mothers

```{r Fig 1, fig.height=5, fig.width=8}
# prevalence for publication - grey scale, no titles, captions

# nb: theme_bw has to come before other theme options
Fig1 <- prev + geom_point(na.rm = F, size=2) + 
  labs(x=NULL, y="% of households with head < 65") + theme_bw() + theme(text=element_text(family="sans"), legend.position = "top", legend.text=element_text(size=12), legend.title=element_blank(), axis.text.x= element_text(angle=45, size=12), axis.text.y=element_text(size=12), axis.title=element_text(size=12), panel.border = element_blank()) + scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(0,0.125)) + scale_color_grey() 
Fig1
ggsave(here("Figures", "Fig1.pdf"), plot=Fig1)
ggsave(here("Figures", "Fig1.png"), plot=Fig1)
```

# Figure 2: Single Parents’ Probability of Non-Employment vs. Couple Families by Gender

```{r Fig 2, echo=FALSE, fig.height=4, fig.width=6}
LIS_emp <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Nonemployment", startRow=2) 
names(LIS_emp) <- make.unique(names(LIS_emp)) # distinguishes model estimates 
View(LIS_emp)

mumemp <- select(LIS_emp, Country, penalty, se) %>%
  mutate(upper=penalty+se, lower=penalty-se, gender="Single Mothers", penalty=penalty) %>% select(Country, penalty, upper, lower, gender) 

dademp <- select(LIS_emp, Country, penalty.1, se.1) %>%
  mutate(upper=penalty.1+se.1, lower=penalty.1-se.1, gender="Single Fathers", penalty=penalty.1) %>% select(Country, penalty, upper, lower, gender) 

LIS_empall <- rbind(mumemp, dademp) 
View(LIS_empall) 

# w/o countries where sample size of single fathers is < 100
samp_size <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Sample_Size") 
emp_samp <- inner_join(LIS_empall, samp_size, by="Country")
View(emp_samp)

LIS_empsub <- emp_samp %>% filter(Single.Fathers > 100) %>% select(Country, penalty, upper, lower, gender) 
View(LIS_empsub)

emp <- ggplot(LIS_empsub, aes(x=reorder(Country, desc(penalty)), y=penalty, ymin=lower, ymax=upper)) # orders by average penalty across models 

Fig2 <- emp + geom_pointrange(aes(color=gender), position=position_dodge(width=0.5)) + labs(x=NULL, y="Probability of Non-Employment \n vs. Couple Families") + theme_bw() + theme(text=element_text(family="sans"), legend.position = "top", legend.text=element_text(size=12), legend.title=element_blank(), axis.text.x= element_text(angle=75, size=12),  axis.title=element_text(size=12), panel.border = element_blank()) +  scale_y_continuous(labels =scales::percent_format(accuracy=1)) + scale_color_manual(values=c("grey", "black")) + geom_hline(yintercept = 0, lty="dashed")
Fig2
ggsave(here("Figures", "Fig2.pdf"))
ggsave(here("Figures", "Fig2.png"))
```

```{r penalties, echo=FALSE, fig.height=9, fig.width=7}
#Figure 3: penalities
LIS_pen <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Penalties", startRow=2) # drops first row with model names 
names(LIS_pen) <- make.unique(names(LIS_pen)) # distinguishes model estimates 
View(LIS_pen)

model0 <- select(LIS_pen, Country, penalty, se) %>%
  mutate(model="BFH", penalty=penalty, upper=penalty, lower=penalty) %>%
  select(Country, penalty, upper, lower, model) 

model1 <- select(LIS_pen, Country, penalty.1, se.1) %>%
  mutate(upper=penalty.1+se.1, lower=penalty.1-se.1, model="BFH, 2010", penalty=penalty.1) %>%
 select(Country, penalty, upper, lower, model)

model2 <- select(LIS_pen, Country, penalty.2, se.2) %>%
  mutate(upper=penalty.2+se.2, lower=penalty.2-se.2, model="Adj.1", penalty=penalty.2) %>%
  select(Country, penalty, upper, lower, model) %>%  filter(!is.na(penalty))

model3 <- select(LIS_pen, Country, penalty.3, se.3) %>%
  mutate(upper=penalty.3+se.3, lower=penalty.3-se.3, model="Adj.2, \n minus multiple earner control", penalty=penalty.3) %>%
  select(Country, penalty, upper, lower, model)

model4 <- select(LIS_pen, Country, penalty.4, se.4) %>%
  mutate(upper=penalty.4+se.4, lower=penalty.4-se.4, model="Adj.3, \n minus nonemployment control", penalty=penalty.4) %>%
  select(Country, penalty, upper, lower, model) 

LIS_pen3 <- rbind(model4, model3, model0) 

LIS_pen3 <- LIS_pen3 %>% mutate(model = fct_relevel(model, "Adj.3, \n minus nonemployment control", "Adj.2, \n minus multiple earner control", "BFH")) %>% arrange(-xtfrm(model), penalty) 
View(LIS_pen3)

pen3 <- ggplot(data=LIS_pen3, aes(x=reorder(Country, penalty), y=penalty, ymin=lower, ymax=upper)) # orders by average penalty across models 

# fig 3, in colour 
penalties <- pen3 + geom_pointrange(aes(color=model), position = position_dodge(width = 1)) + labs(x=NULL, y="Probability of Single Mothers' Poverty \n vs. Couple Families", title="Poverty Penalties to Single Motherhood", caption="Source: Luxembourg Income Study, 2010.") + theme(axis.title=element_text(size=10)) +
scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(-0.1,0.40)) + geom_hline(yintercept = 0, lty="dashed") + coord_flip()
penalties
ggsave(here("Figures", "penalities.pdf"))
```

```{r fig 3, echo=FALSE, fig.width = 7, fig.height = 8}
#fig 3 in greyscale 
Fig3 <- pen3 + geom_pointrange(aes(color=model), position = position_dodge(width = 0.8))  + labs(x=NULL, y="Probability of Poverty for Single Mothers \n vs. Reference Group") + scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(-0.1,0.40)) + scale_color_grey() + coord_flip() + theme_bw() + theme(text=element_text(family="sans"), legend.text=element_text(size=12), axis.text=element_text(size=12),  axis.title=element_text(size=12), panel.border = element_blank()) + geom_hline(yintercept = 0, lty="dashed")
Fig3
ggsave(here("Figures","Fig3.pdf"))
ggsave(here("Figures","Fig3.png"))
```

```{r fig 3b horizontal, fig.width=8, fig.height=5}
# horizontal for presentation
Fig3b <- pen3 + geom_pointrange(aes(color=model), position = position_dodge(width = 0.8))  + labs(x=NULL, y="Probability of Poverty for Single Mothers \n vs. Reference Group") + scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(-0.1,0.40)) + scale_color_grey() + theme_bw() + theme(text=element_text(family="sans"), legend.text=element_text(size=12),legend.position = "top", axis.text=element_text(size=12), axis.text.x=element_text(angle=75), axis.title=element_text(size=12), panel.border = element_blank()) + geom_hline(yintercept = 0, lty="dashed")
Fig3b
ggsave(here("Figures","Fig3b.png"))
```


```{r, echo=FALSE}
# Figure 4: single mothers vs fathers
LIS_dads <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Single_Fathers", startRow=2)  
names(LIS_dads) <- make.unique(names(LIS_dads)) # distinguishes model estimates 
View(LIS_dads)

mothers <- select(LIS_dads, Country, penalty, se) %>%
  mutate(upper=penalty+se, lower=penalty-se, model="Single Mothers", penalty=penalty) %>%
  select(Country, penalty, upper, lower, model) 

fathers <- select(LIS_dads, Country, penalty.1, se.1) %>%
  mutate(upper=penalty.1+se.1, lower=penalty.1-se.1, model="Single Fathers", penalty=penalty.1) %>% select(Country, penalty, upper, lower, model) 

LIS_dadsall <- rbind(fathers, mothers) 

samp_size <- read.xlsx(here("LIS_results", "BFH_data.xlsx"), sheet="Sample_Size") 
dad_sel <- inner_join(LIS_dadsall, samp_size, by="Country")

dad_sel <- dad_sel %>% filter(Single.Fathers > 100) 

means <- dad_sel %>% group_by(model) %>% summarize(mean=mean(penalty)); means

dads <- ggplot(data=dad_sel, aes(x=reorder(Country, desc(penalty)), y=penalty, ymin=lower, ymax=upper)) # orders by average penalty across models 

dad_pen <- dads + geom_pointrange(aes(color=model), position=position_dodge(width=0.5)) + labs(x=NULL, y="Probability of Poverty \n vs. Couple Families", title="Poverty Penalties to Single Parenthood", caption="Source: Luxembourg Income Study, 2010.") + theme_bw() + theme(text=element_text(family="sans"), legend.text=element_text(size=12), legend.title=element_blank(), axis.text=element_text(size=12),  axis.title=element_text(size=12), panel.border = element_blank(), legend.position = "top", axis.text.x=element_text(angle=75)) + scale_y_continuous(labels =scales::percent_format(accuracy=1)) + geom_hline(yintercept = 0, lty="dashed")
dad_pen
ggsave(here("Figures", "dad_pen.pdf"))
```

```{r fig 4, echo=FALSE}
#fig 4 in greyscale 
Fig4 <- dads + geom_pointrange(aes(color=model), position=position_dodge(width=0.5)) + labs(x=NULL, y="Probability of Poverty \n vs. Couple Families") + theme_bw() +  theme(text=element_text(family="sans"), legend.text=element_text(size=12), legend.title=element_blank(), axis.text=element_text(size=12),  axis.title=element_text(size=12), panel.border = element_blank(), legend.position = "top", axis.text.x=element_text(angle=75)) + scale_y_continuous(labels =scales::percent_format(accuracy=1), limits=c(-0.25, 0.52)) + scale_color_manual(values=c("grey", "black")) + geom_hline(yintercept = 0, lty="dashed")
Fig4
ggsave(here("Figures","Fig4.pdf"))
ggsave(here("Figures","Fig4.png"))
```


